Comparison of particle trajectories and collision operators for collisional transport in nonaxisymmetric plasmas

### 2014-04-03 doc/sfincsPaper/sfincsPaper.tex

In this work, we examine the validity of several common simplifying assumptions used in numerical neoclassical calculations
for nonaxisymmetric plasmas,
\changed{
both by using a new continuum drift-kinetic code
and by considering analytic properties of the kinetic equation.}
First, neoclassical phenomena are computed for the LHD and W7-X stellarators using several versions of the drift-kinetic
equation, 
including the commonly used incompressible-$\vect{E}\times\vect{B}$-drift approximation
and two other variants,
corresponding to different effective particle trajectories.
It is found that
for electric fields below roughly one third of the resonant value, the different formulations give
nearly identical results, demonstrating the incompressible $\vect{E}\times\vect{B}$-drift approximation
is quite accurate in this regime. 
However, near the electric field resonance,
the models yield substantially different results.
We also compare results for various collision operators, including the full linearized Fokker-Planck operator.
At low collisionality, the radial transport driven by radial
gradients is nearly identical for the different operators,
while in other cases it is found to be important that collisions conserve momentum.
%At high collisionality, it seems necessary to use the full Fokker-Planck operator to obtain all transport coefficients correctly.

@@

In the following section, we motivate the form of the kinetic equation solved by SFINCS,
and detail the three models for particle trajectories that will be compared.
For several of the particle trajectory models,
additional sources/sinks and constraints must be included in the system of equations
for the equations to be well posed and for the solutions to be well behaved.
These issues are explored in section \ref{sec:sources}.
\changed{In section \ref{sec:momentum}, we discuss some observations regarding momentum conservation,
and demonstrate that the electric field terms in the kinetic description
correspond to a component of gyroviscosity in a fluid description
only for the most accurate trajectory model.}
Details of the numerical implementation are given in section \ref{sec:numerics}.
Some of the numerical results presented are given in terms of a transport
matrix, which is defined in section \ref{sec:transportMatrix}.
The numerical results are presented in sections \ref{sec:ErComparison} and \ref{sec:collisionComparison},
in which we discuss the transport matrix elements for the geometries of the LHD and W7-X
stellarators, comparing a variety of assumptions about the particle trajectories and collision operator.
In section \ref{sec:conclusions} we discuss the results and conclude.

@@

Unfortunately, (\ref{eq:kinetic1}) does not contain the physics of $\vect{E}\times\vect{B}$ precession,
since the characteristic curves of this equation correspond only to motion along the magnetic field lines.
Consequently, important transport regimes such as the $\sqrt{\nu}$ regime cannot be obtained
using (\ref{eq:kinetic1}).
To retain $\vect{E}\times\vect{B}$ precession, we also keep the term $(\vE+\vma)\cdot\nabla f_{a1}$
in (\ref{eq:kinetic1}), even though according to the formal ordering it should
appear at next order.  A similar step is made in other stellarator neoclassical
calculations \cite{DKES1, DKES2}.
The mathematical reason why this term is important at low collisionality is that it has different symmetry properties than other, possible larger, terms in (\ref{eq:kinetic1}). For instance, it 
survives if a bounce average is used to annihilate the first term.
(We will not bounce average the kinetic equation here,
but when the collisionality is low, the solution of the full equation becomes
asymptotically close to the solution of the bounce-averaged equation.)

@@

For numerical computations, it is convenient to use coordinates for which the ranges of allowed values
are independent of the other coordinates.  As $W_{a0}$ and $\mu$ do not have this property, it is convenient
to switch to coordinates $x_a$ and $\xi = \vpar/v$.  Carrying out this change of variables on the first term
of (\ref{eq:kinetic3}), we find
\begin{equation}
\dot{\vect{r}} \cdot\left(\nabla f_{a1}\right)_{W_{a0},\mu}
=
\dot{\vect{r}}\cdot \left(\nabla f_{a1}\right)_{x_a,\xi}
+\dot{x}_a \left(\frac{\partial f_{a1}}{\partial x_a}\right)_{\vect{r},\xi}
+\dot{\xi}_a \left(\frac{\partial f_{a1}}{\partial \xi}\right)_{\vect{r},x_a},
\end{equation}
where $\vect{r}$ denotes the position vector,
\begin{equation}
\dot{\vect{r}} = \vpar \vect{b} + \vEo + \vma,
\label{eq:rdot1}
\end{equation}
\begin{equation}
\dot{x}_a = (\vma\cdot\nabla\psi) \left(-\frac{x_a}{2T_a}\frac{dT_a}{d\psi} - \frac{Z_a e}{2T_a x_a}\frac{d\Phi_0}{d\psi}\right),
\label{eq:xdot1}
\end{equation}
and
\begin{equation}
\dot{\xi}_a = -\frac{1-\xi^2}{2 B \xi} \vpar \vect{b}\cdot\nabla B
+\xi(1-\xi^2) \frac{c}{2 B^3} \frac{d\Phi_0}{d\psi}\vect{B}\times\nabla\psi\cdot\nabla B
-\frac{1-\xi^2}{2 B \xi} \vma\cdot\nabla B
.
\label{eq:xidot1}
\end{equation}
For the rest of this work, we will neglect the $\vma$ term in (\ref{eq:rdot1}), the $dT_a/d\psi$ term
in (\ref{eq:xdot1}), and the $\vma\cdot\nabla B$ term in (\ref{eq:xidot1}),
for several reasons.
First, if the  $\vma$ term in (\ref{eq:rdot1}) was retained, we would need to solve a 5D rather
than 4D problem due to the radial coupling (i.e. $\psi$ appearing
as a derivative rather than merely as a parameter). Second,
once radial coupling is dropped, we must also drop the $dT_a/d\psi$ term
in (\ref{eq:xdot1}) and the $\vma\cdot\nabla B$ term in (\ref{eq:xidot1})
in order for $\mu$ to be conserved.
Third, dropping these terms conveniently eliminates all dependence of the transport matrix (defined in section \ref{sec:transportMatrix})
on $d T_a/d\psi$, $dB/d\psi$,
\changed{
and $\rho_*$.
Fourth, dropping these terms amounts to taking the limit $\rho_* \to 0$
(while keeping the $d\Phi_0/d\psi$ terms finite), 
and this limit is already complicated and interesting to explore without the extra complexity
of finite-$\rho_*$ corrections.
Fifth, 
we wish to focus on the effects of the radial electric field.
The omitted terms may be important in other situations, but here our primary interest is the treatment of the $d\Phi_0/d\psi$ terms.
Finally,
these omitted terms would significantly complicate the analysis in section \ref{sec:sources},
in which we will examine moments of the kinetic equation.
}

@@

\changed{
Note that for both the DKES and full trajectories, the left-hand side of the
kinetic equation (\ref{eq:kinetic4})
can be written in the conservative form
\begin{equation}
\frac{1}{J}\left[
\nabla\cdot\left( J \dot{\vect{r}}_a f_{a1} \right) 
+ \frac{\partial}{\partial \xi}\left( J \dot{\xi}_a f_{a1} \right)
+ \frac{\partial}{\partial x_a} \left( J \dot{x}_a f_{a1} \right)
\right] - C_a
\label{eq:conservativeForm}
\end{equation}
where $J=x_a^2$ is the Jacobian of the transformation between Cartesian velocity coordinates
and the coordinates $x_a$, $\xi$, and gyrophase. However, for the partial trajectories,
the left-hand side of (\ref{eq:kinetic4}) is not equivalent
to (\ref{eq:conservativeForm}).
}

@@

If one attempts to solve the kinetic equation (\ref{eq:kinetic4}) numerically using either the full or partial trajectories
and $E_r \neq 0$, unphysical results will be obtained, with
the numerical solution not converging as resolution parameters are increased.
We now explore the reason for this behavior.
We will then describe a modified form of the kinetic equation which robustly produces more sensible results.
\changed{The issues discussed in this section are related to moment equations for mass and energy; momentum has a different status
and will be examined in the next section.}

@@

One motivation for use of the DKES trajectory model is now apparent:
it is the only model that avoids the imposition of one or more unphysical
constraints on the distribution function when $d\Phi_0/d\psi \ne 0$,
constraints which cause an $E_r \ne 0$ calculation to be a singular perturbation
of an $E_r = 0$ calculation.

@@

\changed{
This system of sources and constraints solves not only the problem described above when $E_r \ne 0$,
but also a different problem that remains even when $E_r=0$ and/or when the DKES trajectories are used:
the kinetic equation has a null space.
If the conditions $\left< \int d^3v \,f_{a1}\right>=0$ and $\left< \int d^3v\, v^2 f_{a1}\right>=0$
were not imposed, any linear combination of $F_{a}$ and $F_{a} v^2$ could be added to one solution
of the kinetic equation to obtain another solution. Upon discretization, one would obtain a non-invertible
(or at least very poorly conditioned) linear system, but the imposition of these two extra constraints
makes the system of equations invertible. 
}

\changed{
Such is the case when the full linearized Fokker-Planck 
collision operator is used, but the situation is different when either the pitch-angle scattering
operator or momentum-conserving model operator are used instead, for then the kinetic equation has a larger null space:
any function of $v$ is then a homogeneous solution of the kinetic equation.
As the dimension of the null space is then equal to $N_x$ (the number of grid points in $x_a$) rather than
2, it takes $N_x$ rather than 2 constraint equations to eliminate the null space for these collision operators.
We choose these $N_x$ constraints to be $\left< \int_{-1}^1 d\xi\, f_{a1}\right>=0$ (imposed at each grid point in $x_a$.)
To keep the linear system square, we must then have $N_x$ rather than 2 unknowns related to the sources. 
This is accomplished by letting the source be a general function of $x_a$ instead
of (\ref{eq:sources}) when either the pitch-angle scattering or momentum-conserving model collision operator are used.
This alternative system of $N_x$ sources and constraints is an equally reasonable solution to the earlier
conservation problem.
}

\changed{
To summarize, the sources and extra constraint equations serve two independent purposes.
First, when $E_r \ne 0$, the sources are needed to eliminate the conservation problems, and the extra constraints then keep the linear system square (number of equations = number of unknowns) upon discretization.
Second, even when $E_r = 0$, and even for the DKES trajectories in which sources are not required, the constraints are needed to eliminate the null space in the kinetic equation, and the source terms are a convenient way to keep the linear system square upon discretization.
The first problem can be solved with either the source (\ref{eq:sources}) or a general speed-dependent source $S_a(x_a)$.
However, to solve the second problem, the number of constraints should match the dimensionality of the null space.
For this reason, we apply the source (\ref{eq:sources}) with 2 constraints when the Fokker-Planck operator is used,
while we apply the general speed-dependent source $S_a(x_a)$ with $N_x$ constraints when either of the other two collision operators is used.
}

\begin{comment}
The aforemention mass and energy sources and constraints work when the Fokker-Planck collision operator
is used, 
but not when the pure pitch-angle scattering or momentum-conserving model collision operator is used, since in these cases the kinetic equation has a larger null space
(any function of $v$ is a solution.)
For these collision operators, we instead allow a general source $S_a(\psi,v)$,
introducing $N_x$ unknowns on each flux surface, where $N_x$ is the number of grid points in $v$.
To supply an equal number of constraints, we impose the equations $\left< \int_{-1}^1 d\xi\, f_{a1}\right>=0$
at each grid point in $v$.
\end{comment}

@@

\changed{
Parallel momentum has a different status to density and energy, in that density and energy are conserved by
the collisionless motion while parallel momentum is not, due to the mirror force.
(For example, considering the simple case of a single ion species with $d\Phi_0/d\psi=0$, $F_i$ and $v^2 F_i$ are homogeneous
solutions to the kinetic equation, whereas $v_{||} F_i$ is not.)
A consequence is that there does not appear to be a false constraint for $E_r \ne 0$ arising
from the $\left< \int d^3 v\, v_{||} (\ldots)\right>$ moment of the various forms of the kinetic equation,
i.e. there is no exact analogue to (\ref{eq:unphysicalMassConservation}), (\ref{eq:unphysicalEnergyConservation1}),
or (\ref{eq:unphysicalEnergyConservation2}) for momentum.
When the momentum moment of the various forms of the drift-kinetic equation is taken,
even if a factor of $B$ or $1/B$ is included in the flux surface average, 
a collisionless term remains that is not proportional to $d\Phi_0/d\psi$.
Consequently, for all the trajectory models,
$d\Phi_0/d\psi=0$ is a well behaved rather than singular
limit of the momentum moment equation.
}

\changed{
Nonetheless, it is interesting to compare the $\int d^3v\, m_a v_{||} (\ldots)$ moment equations
for each drift-kinetic trajectory model to the full parallel momentum fluid equation. This later equation,
the moment of the full Fokker-Planck equation, is
\begin{equation}
0 = -\vect{b}\cdot\left( \nabla\cdot\tens{\Pi}_a\right)
+ Z_a e n_a E_{||} + F_{a||}
\label{eq:fluidParallelMomentum}
\end{equation}
where $\tens{\Pi_a} = m_a \int d^3v\, f_a \vect{v}\vect{v}$ is the stress tensor
and $F_{a||}$ is the parallel component of friction.
First, consider the case of no radial electric field.
Recalling
$f_a = F_a(\psi)\left[ 1 - Z_a e \Phi_1/T_a\right] + f_{a1}$, 
the stress tensor is given to the accuracy needed by 
$\tens{\Pi}_a \approx p_a(\psi)\left[1 - Z_a e \Phi_1/T_a\right] \tens{I} + \tens{\Pi}_{a1}$
where $\tens{\Pi}_{a1} = p_{a 1 \bot} \tens{I} + (p_{a1||} - p_{a1\bot}) \vect{b}\vect{b}$,
$p_{a1||} = m_a \int d^3v\, f_{a1} v_{||}^2$, and $p_{a1\bot} = m_a \int d^3v\, f_{a1} v_{\bot}^2/2$.
Notice the $\int d^3v\, m_a v_{||} (\ldots)$ moment of the streaming and mirror terms in (\ref{eq:kinetic4})-(\ref{eq:fullTrajectories}) is
\begin{eqnarray}
\int d^3v\, m_a v_{||} \left[ v_{||} \vect{b}\cdot\nabla f_{a1} - \frac{1-\xi^2}{2B}v(\vect{b}\cdot\nabla B) \frac{\partial f_{a1}}{\partial \xi} \right]
& = &
\vect{b}\cdot\nabla p_{a1||} + \frac{ p_{a1\bot} - p_{a1 ||}}{B} \vect{b}\cdot\nabla B \nonumber \\
& = &
\vect{b} \cdot \left(\nabla \cdot \tens{\Pi}_{a1} \right).
\end{eqnarray}
Using this result with (\ref{eq:gauge}), the $m_a v_{||}$ moment of 
the drift-kinetic equation (\ref{eq:kinetic4})-(\ref{eq:fullTrajectories}) matches
the full fluid parallel momentum equation (\ref{eq:fluidParallelMomentum})
at least when $E_r=0$.
}

\changed{
Now consider how the situation changes when a radial electric field is introduced. We first
compute the change to the fluid parallel momentum equation caused by a new contribution
to the viscosity.
Examining the $\left( \vect{E} + c^{-1} \vect{v}\times\vect{B}\right) \cdot \nabla_{\vect{v}} f_a$ terms in 
the full Fokker-Planck equation and integrating in gyrophase, one sees the gyrophase-dependent part of the 
distribution function $\tilde{f}_a$ will include the following terms proportional to the electric field:
\begin{equation}
\tilde{f}_{aE} = \frac{c}{B} \vect{v}\cdot\vect{b}\times\vect{E} \left[ \frac{1}{v} \frac{\partial \bar{f}_a}{\partial v}
-\frac{\xi}{v^2} \frac{\partial \bar{f}_a}{\partial \xi} \right]
\label{eq:fTilde}
\end{equation}
as reflected (using different independent variables) in eq (17) of Ref. \cite{Hazeltine} and eq (6) of Ref. \cite{SimakovCattoSecondOrderDKE}.
Here, $\bar{f}_a$ is the gyrophase-independent part of the distribution function.
The associated contribution $\tens{\Pi}_{aE} = m_a \int d^3v\, \tilde{f}_{aE}\vect{v}\vect{v}$
to the pressure tensor is calculated in equations (27)-(36) of Ref. \cite{SimakovCattoSecondOrderDKE},
with the result
\begin{equation}
\tens{\Pi}_{aE} = m_a n_a V_{a||} \left( \vect{b}\vect{v}_E + \vect{v}_E \vect{b} \right)
\label{eq:PiE}
\end{equation}
where $n_a V_{a||} = \int d^3v\, f_a v_{||}$. Note that this contribution to the stress tensor is a part of the gyroviscosity and is off-diagonal.
Using $\vect{E} \approx -\nabla \Phi_0(\psi)$ and $\vect{B} \cdot \left\{ \nabla \cdot \left[ \vect{B}\vect{B}\times\nabla\psi + (\vect{B}\times\nabla\psi) \vect{B} \right] \right\}
=2B^2\nabla\psi\cdot\nabla\times\vect{B}=0$,
we then find the contribution to the parallel momentum equation (\ref{eq:fluidParallelMomentum}) from (\ref{eq:fTilde})-(\ref{eq:PiE}) is
\begin{equation}
\vect{b}\cdot(\nabla\cdot\tens{\Pi}_{aE}) = 
c m_a B (d\Phi_0/d\psi)
\vect{B}\times\nabla\psi\cdot\nabla\left(n_a V_{a||}/B^3\right).
\label{eq:PiEParallel}
\end{equation}
}

>>>>

In this work, we examine the validity of several common simplifying assumptions used in numerical neoclassical calculations
for nonaxisymmetric plasmas,
both by using a new continuum drift-kinetic code
and by considering analytic properties of the kinetic equation.
First, neoclassical phenomena are computed for the LHD and W7-X stellarators using several versions of the drift-kinetic
equation, 
including the commonly used incompressible-$\vect{E}\times\vect{B}$-drift approximation
and two other variants,
corresponding to different effective particle trajectories.
It is found that
for electric fields below roughly one third of the resonant value, the different formulations give
nearly identical results, demonstrating the incompressible $\vect{E}\times\vect{B}$-drift approximation
is quite accurate in this regime. 
However, near the electric field resonance,
the models yield substantially different results.
We also compare results for various collision operators, including the full linearized Fokker-Planck operator.
At low collisionality, the radial transport driven by radial
gradients is nearly identical for the different operators,
while in other cases it is found to be important that collisions conserve momentum.
%At high collisionality, it seems necessary to use the full Fokker-Planck operator to obtain all transport coefficients correctly.

@@

In the following section, we motivate the form of the kinetic equation solved by SFINCS,
and detail the three models for particle trajectories that will be compared.
For several of the particle trajectory models,
additional sources/sinks and constraints must be included in the system of equations
for the equations to be well posed and for the solutions to be well behaved.
These issues are explored in section \ref{sec:sources}.
In section \ref{sec:momentum}, we discuss some observations regarding momentum conservation,
and demonstrate that the electric field terms in the kinetic description
correspond to a component of gyroviscosity in a fluid description
only for the most accurate trajectory model.
Details of the numerical implementation are given in section \ref{sec:numerics}.
Some of the numerical results presented are given in terms of a transport
matrix, which is defined in section \ref{sec:transportMatrix}.
The numerical results are presented in sections \ref{sec:ErComparison} and \ref{sec:collisionComparison},
in which we discuss the transport matrix elements for the geometries of the LHD and W7-X
stellarators, comparing a variety of assumptions about the particle trajectories and collision operator.
In section \ref{sec:conclusions} we discuss the results and conclude.

@@

Unfortunately, (\ref{eq:kinetic1}) does not contain the physics of $\vect{E}\times\vect{B}$ precession,
since the characteristic curves of this equation correspond only to motion along the magnetic field lines.
Consequently, important transport regimes such as the $\sqrt{\nu}$ regime cannot be obtained
using (\ref{eq:kinetic1}).
To retain $\vect{E}\times\vect{B}$ precession, we also keep the term $(\vE+\vma)\cdot\nabla f_{a1}$
in (\ref{eq:kinetic1}), even though according to the formal ordering it should
appear at next order.  A similar step is made in other stellarator neoclassical
calculations \cite{DKES1, DKES2}.
The mathematical reason why this term is important at low collisionality is that it has different symmetry properties than other, possibly larger, terms in (\ref{eq:kinetic1}). For instance, it 
survives if a bounce average is used to annihilate the first term.
(We will not bounce average the kinetic equation here,
but when the collisionality is low, the solution of the full equation becomes
asymptotically close to the solution of the bounce-averaged equation.)

@@

For numerical computations, it is convenient to use coordinates for which the ranges of allowed values
are independent of the other coordinates.  As $W_{a0}$ and $\mu$ do not have this property, it is convenient
to switch to coordinates $x_a$ and $\xi = \vpar/v$.  Carrying out this change of variables on the first term
of (\ref{eq:kinetic3}), we find
\begin{equation}
\dot{\vect{r}} \cdot\left(\nabla f_{a1}\right)_{W_{a0},\mu}
=
\dot{\vect{r}}\cdot \left(\nabla f_{a1}\right)_{x_a,\xi}
+\dot{x}_a \left(\frac{\partial f_{a1}}{\partial x_a}\right)_{\vect{r},\xi}
+\dot{\xi}_a \left(\frac{\partial f_{a1}}{\partial \xi}\right)_{\vect{r},x_a},
\end{equation}
where $\vect{r}$ denotes the position vector,
\begin{equation}
\dot{\vect{r}} = \vpar \vect{b} + \vEo + \vma,
\label{eq:rdot1}
\end{equation}
\begin{equation}
\dot{x}_a = (\vma\cdot\nabla\psi) \left(-\frac{x_a}{2T_a}\frac{dT_a}{d\psi} - \frac{Z_a e}{2T_a x_a}\frac{d\Phi_0}{d\psi}\right),
\label{eq:xdot1}
\end{equation}
and
\begin{equation}
\dot{\xi}_a = -\frac{1-\xi^2}{2 B \xi} \vpar \vect{b}\cdot\nabla B
+\xi(1-\xi^2) \frac{c}{2 B^3} \frac{d\Phi_0}{d\psi}\vect{B}\times\nabla\psi\cdot\nabla B
-\frac{1-\xi^2}{2 B \xi} \vma\cdot\nabla B
.
\label{eq:xidot1}
\end{equation}
For the rest of this work, we will neglect the $\vma$ term in (\ref{eq:rdot1}), the $dT_a/d\psi$ term
in (\ref{eq:xdot1}), and the $\vma\cdot\nabla B$ term in (\ref{eq:xidot1}),
for several reasons.
First, if the  $\vma$ term in (\ref{eq:rdot1}) was retained, we would need to solve a 5D rather
than 4D problem due to the radial coupling (i.e. $\psi$ appearing
as a derivative rather than merely as a parameter). Second,
once radial coupling is dropped, we must also drop the $dT_a/d\psi$ term
in (\ref{eq:xdot1}) and the $\vma\cdot\nabla B$ term in (\ref{eq:xidot1})
in order for $\mu$ to be conserved.
Third, dropping these terms conveniently eliminates all dependence of the transport matrix (defined in section \ref{sec:transportMatrix})
on $d T_a/d\psi$, $dB/d\psi$,
and $\rho_*$.
Fourth, dropping these terms amounts to taking the limit $\rho_* \to 0$
(while keeping the $d\Phi_0/d\psi$ terms finite), 
and this limit is already complicated and interesting to explore without the extra complexity
of finite-$\rho_*$ corrections.
Fifth, 
we wish to focus on the effects of the radial electric field.
The omitted terms may be important in other situations, but here our primary interest is the treatment of the $d\Phi_0/d\psi$ terms.
Finally,
these omitted terms would significantly complicate the analysis in section \ref{sec:sources},
in which we will examine moments of the kinetic equation.

@@

Note that for both the DKES and full trajectories, the left-hand side of the
kinetic equation (\ref{eq:kinetic4})
can be written in the conservative form
\begin{equation}
\frac{1}{J}\left[
\nabla\cdot\left( J \dot{\vect{r}}_a f_{a1} \right) 
+ \frac{\partial}{\partial \xi}\left( J \dot{\xi}_a f_{a1} \right)
+ \frac{\partial}{\partial x_a} \left( J \dot{x}_a f_{a1} \right)
\right] - C_a
\label{eq:conservativeForm}
\end{equation}
where $J=x_a^2$ is the Jacobian of the transformation between Cartesian velocity coordinates
and the coordinates $x_a$, $\xi$, and gyrophase. However, for the partial trajectories,
the left-hand side of (\ref{eq:kinetic4}) is not equivalent
to (\ref{eq:conservativeForm}).

@@

If one attempts to solve the kinetic equation (\ref{eq:kinetic4}) numerically using either the full or partial trajectories
and $E_r \neq 0$, unphysical results will be obtained, with
the numerical solution not converging as resolution parameters are increased.
We now explore the reason for this behavior.
We will then describe a modified form of the kinetic equation which robustly produces more sensible results.
The issues discussed in this section are related to moment equations for mass and energy; momentum has a different status
and will be examined in the next section.

@@

One motivation for use of the DKES trajectory model is now apparent:
it is the only model (of the three condered here) that avoids the imposition of one or more unphysical
constraints on the distribution function when $d\Phi_0/d\psi \ne 0$,
constraints which cause an $E_r \ne 0$ calculation to be a singular perturbation
of an $E_r = 0$ calculation.

@@

This system of sources and constraints solves not only the problem described above when $E_r \ne 0$,
but also a different problem that remains even when $E_r=0$ and/or when the DKES trajectories are used:
the kinetic equation has a null space.
If the conditions $\left< \int d^3v \,f_{a1}\right>=0$ and $\left< \int d^3v\, v^2 f_{a1}\right>=0$
were not imposed, any linear combination of $F_{a}$ and $F_{a} v^2$ could be added to one solution
of the kinetic equation to obtain another solution. Upon discretization, one would obtain a non-invertible
(or at least very poorly conditioned) linear system, but the imposition of these two extra constraints
makes the system of equations invertible.

Such is the case when the full linearized Fokker-Planck 
collision operator is used, but the situation is different when either the pitch-angle scattering
operator or momentum-conserving model operator are used instead, for then the kinetic equation has a larger null space:
any function of $v$ is then a homogeneous solution of the kinetic equation.
As the dimension of the null space is then equal to $N_x$ (the number of grid points in $x_a$) rather than
2, it takes $N_x$ rather than 2 constraint equations to eliminate the null space for these collision operators.
We choose these $N_x$ constraints to be $\left< \int_{-1}^1 d\xi\, f_{a1}\right>=0$ (imposed at each grid point in $x_a$.)
To keep the linear system square, we must then have $N_x$ rather than 2 unknowns related to the sources. 
This is accomplished by letting the source be a general function of $x_a$ instead
of (\ref{eq:sources}) when either the pitch-angle scattering or momentum-conserving model collision operator are used.
This alternative system of $N_x$ sources and constraints is an equally reasonable solution to the earlier
conservation problem.

To summarize, the sources and extra constraint equations serve two independent purposes.
First, when $E_r \ne 0$, the sources are needed to eliminate the conservation problems, and the extra constraints then keep the linear system square (number of equations = number of unknowns) upon discretization.
Second, even when $E_r = 0$, and even for the DKES trajectories in which sources are not required, the constraints are needed to eliminate the null space in the kinetic equation, and the source terms are a convenient way to keep the linear system square upon discretization.
The first problem can be solved with either the source (\ref{eq:sources}) or a general speed-dependent source $S_a(x_a)$.
However, to solve the second problem, the number of constraints should match the dimensionality of the null space.
For this reason, we apply the source (\ref{eq:sources}) with 2 constraints when the Fokker-Planck operator is used,
while we apply the general speed-dependent source $S_a(x_a)$ with $N_x$ constraints when either of the other two collision operators is used.

@@

Parallel momentum has a different status to density and energy, in that density and energy are conserved by
the collisionless motion while parallel momentum is not, due to the mirror force.
(For example, considering the case of a single ion species with $d\Phi_0/d\psi=0$, $F_i$ and $v^2 F_i$ are homogeneous
solutions to the kinetic equation, whereas $v_{||} F_i$ is not.)
A consequence is that there does not appear to be a false constraint for $E_r \ne 0$ arising
from the $\left< \int d^3 v\, v_{||} (\ldots)\right>$ moment of the various forms of the kinetic equation,
i.e. there is no analogue to (\ref{eq:unphysicalMassConservation}), (\ref{eq:unphysicalEnergyConservation1}),
or (\ref{eq:unphysicalEnergyConservation2}) for momentum.
When the momentum moment of the various forms of the drift-kinetic equation is taken,
even if a factor of $B$ or $1/B$ is included in the flux surface average, 
a collisionless term remains that is not proportional to $d\Phi_0/d\psi$.
Consequently, for all the trajectory models,
$d\Phi_0/d\psi=0$ is a well behaved rather than singular
limit of the momentum moment equation.

Nonetheless, it is interesting to compare the $\int d^3v\, m_a v_{||} (\ldots)$ moment equations
for each drift-kinetic trajectory model to the full parallel momentum fluid equation. This later equation,
the moment of the full Fokker-Planck equation, is
\begin{equation}
0 = -\vect{b}\cdot\left( \nabla\cdot\tens{\Pi}_a\right)
+ Z_a e n_a E_{||} + F_{a||}
\label{eq:fluidParallelMomentum}
\end{equation}
where $\tens{\Pi_a} = m_a \int d^3v\, f_a \vect{v}\vect{v}$ is the stress tensor
and $F_{a||}$ is the parallel component of friction.
First, consider the case of no radial electric field.
Recalling
$f_a = F_a(\psi)\left[ 1 - Z_a e \Phi_1/T_a\right] + f_{a1}$, 
the stress tensor is given to the accuracy needed by 
$\tens{\Pi}_a \approx p_a(\psi)\left[1 - Z_a e \Phi_1/T_a\right] \tens{I} + \tens{\Pi}_{a1}$
where $\tens{\Pi}_{a1} = p_{a 1 \bot} \tens{I} + (p_{a1||} - p_{a1\bot}) \vect{b}\vect{b}$,
$p_{a1||} = m_a \int d^3v\, f_{a1} v_{||}^2$, and $p_{a1\bot} = m_a \int d^3v\, f_{a1} v_{\bot}^2/2$.
Notice the $\int d^3v\, m_a v_{||} (\ldots)$ moment of the streaming and mirror terms in (\ref{eq:kinetic4})-(\ref{eq:fullTrajectories}) is
\begin{eqnarray}
\int d^3v\, m_a v_{||} \left[ v_{||} \vect{b}\cdot\nabla f_{a1} - \frac{1-\xi^2}{2B}v(\vect{b}\cdot\nabla B) \frac{\partial f_{a1}}{\partial \xi} \right]
& = &
\vect{b}\cdot\nabla p_{a1||} + \frac{ p_{a1\bot} - p_{a1 ||}}{B} \vect{b}\cdot\nabla B \nonumber \\
& = &
\vect{b} \cdot \left(\nabla \cdot \tens{\Pi}_{a1} \right).
\end{eqnarray}
Using this result with (\ref{eq:gauge}), the $m_a v_{||}$ moment of 
the drift-kinetic equation (\ref{eq:kinetic4})-(\ref{eq:fullTrajectories}) matches
the full fluid parallel momentum equation (\ref{eq:fluidParallelMomentum})
at least when $E_r=0$.

Now consider how the situation changes when a radial electric field is introduced. We first
compute the change to the fluid parallel momentum equation caused by a new contribution
to the viscosity.
Examining the $\left( \vect{E} + c^{-1} \vect{v}\times\vect{B}\right) \cdot \nabla_{\vect{v}} f_a$ terms in 
the full Fokker-Planck equation and integrating in gyrophase, one sees the gyrophase-dependent part of the 
distribution function $\tilde{f}_a$ will include the following terms proportional to the electric field:
\begin{equation}
\tilde{f}_{aE} = \frac{c}{B} \vect{v}\cdot\vect{b}\times\vect{E} \left[ \frac{1}{v} \frac{\partial \bar{f}_a}{\partial v}
-\frac{\xi}{v^2} \frac{\partial \bar{f}_a}{\partial \xi} \right]
\label{eq:fTilde}
\end{equation}
as reflected (using different independent variables) in eq (17) of Ref. \cite{Hazeltine} and eq (6) of Ref. \cite{SimakovCattoSecondOrderDKE}.
Here, $\bar{f}_a$ is the gyrophase-independent part of the distribution function.
The associated contribution $\tens{\Pi}_{aE} = m_a \int d^3v\, \tilde{f}_{aE}\vect{v}\vect{v}$
to the pressure tensor is calculated in equations (27)-(36) of Ref. \cite{SimakovCattoSecondOrderDKE},
with the result
\begin{equation}
\tens{\Pi}_{aE} = m_a n_a V_{a||} \left( \vect{b}\vect{v}_E + \vect{v}_E \vect{b} \right)
\label{eq:PiE}
\end{equation}
where $n_a V_{a||} = \int d^3v\, f_a v_{||}$. Note that this contribution to the stress tensor is a part of the gyroviscosity and is off-diagonal.
Using $\vect{E} \approx -\nabla \Phi_0(\psi)$ and $\vect{B} \cdot \left\{ \nabla \cdot \left[ \vect{B}\vect{B}\times\nabla\psi + (\vect{B}\times\nabla\psi) \vect{B} \right] \right\}
=2B^2\nabla\psi\cdot\nabla\times\vect{B}=0$,
we then find the contribution to the parallel momentum equation (\ref{eq:fluidParallelMomentum}) from (\ref{eq:fTilde})-(\ref{eq:PiE}) is
\begin{equation}
\vect{b}\cdot(\nabla\cdot\tens{\Pi}_{aE}) = 
c m_a B (d\Phi_0/d\psi)
\vect{B}\times\nabla\psi\cdot\nabla\left(n_a V_{a||}/B^3\right).
\label{eq:PiEParallel}
\end{equation}

For comparison, let us consider the $\int d^3v\, m_a v_{||} (\ldots)$ moment of the radial electric field
terms in the drift-kinetic equation for various trajectory models, to see if the results agree with (\ref{eq:PiEParallel}).
For the full trajectories, the moment of the $d\Phi_0/d\psi$ terms in (\ref{eq:kinetic4})-(\ref{eq:fullTrajectories}) is
\begin{eqnarray}
\int d^3v\, m_a v_{||} \left[ 
\frac{c}{B^2} \frac{d\Phi_0}{d\psi} \vect{B}\times\nabla\psi\cdot\nabla f_{a1}
- (\vma \cdot\nabla\psi) \frac{Z_a e}{2 T_a x_a} \frac{d\Phi_0}{d\psi} \frac{\partial f_{a1}}{\partial x_a}
\right.
\nonumber \\
\left.
+ \xi(1-\xi^2) \frac{c}{2B^3} \frac{d\Phi_0}{d\psi} (\vect{B}\times\nabla\psi\cdot\nabla B)\frac{\partial f_{a}}{\partial \xi}
\right]
\nonumber \\
= 
c m_a B (d\Phi_0/d\psi)
\vect{B}\times\nabla\psi\cdot\nabla\left(n_a V_{a||}/B^3\right),
\label{eq:fullViscosity}
\end{eqnarray}
obtained by integrating by parts in $x_a$ and $\xi$.
Thus, the full-trajectory model agrees with the full fluid parallel momentum equation: (\ref{eq:fullViscosity}) = (\ref{eq:PiEParallel}).
However, this agreement is
not shared by the DKES model: the moment of the $d\Phi_0/d\psi$ term in (\ref{eq:DKESTrajectories}) is
\begin{eqnarray}
\int d^3v\, m_a v_{||} \left[ 
\frac{c}{\left< B^2 \right> } \frac{d\Phi_0}{d\psi} \vect{B}\times\nabla\psi\cdot\nabla f_{a1}
\right]
= 
\frac{c m_a}{ \left< B^2 \right>}
\frac{d\Phi_0}{d\psi}
\vect{B}\times\nabla\psi\cdot\nabla (n_a V_{a||}),
\label{eq:DKESViscosity}
\end{eqnarray}
which does not equal (\ref{eq:PiEParallel}). The corresponding result for the partial trajectories,
obtained by replacing $\left< B^2 \right> \to B^2 $ in (\ref{eq:DKESViscosity}), also does not match
(\ref{eq:PiEParallel}). Thus, the DKES and partial trajectory models do not correctly account for the parallel
viscous force as the full trajectory model does.
